---
title: Linear Regression
author: George Whittington
date: today
date-format: long
---

## Load Libraries

```{r}
library(MASS)
library(tidyverse)
```

## Load Data

```{r}
df <- Boston
head(df)
```

## Scatter Plot

### Prof Plot
```{r}
plot(x = Boston$lstat, y = Boston$medv,
     xlab = "x (lstat)", ylab = "y (medv)",
     main = "Scatter plot of x and y",
     col="red", pch = 20, cex.lab = 1.5, cex.main = 1.5
)
```

### My Plot

```{r}
df |> 
	ggplot(aes(lstat, medv)) +
	geom_point(color="red") +
	labs(
		title = "Scatter plot of x and y"
	)
```

## Simple Linear Model

### R Function

```{r}
simple_model <- lm(medv ~ lstat, data=df)
summary(simple_model)
```

```{r}
anova(simple_model)
```

### Manual

#### Calculation of $\beta_1$ and $\beta_0$

```{r}
# sample size
n <- nrow(df)

# x variable
x <- df$lstat
# y variable
y <- df$medv

# mean of x
x_bar <- mean(x)
# mean of y
y_bar <- mean(y)

# variance of x: var(x)
var_x <- sum((x - x_bar)^2) / (n - 1)
# variance of y: var(y)
var_y <- sum((y - y_bar)^2) / (n - 1)
# standard deviation of x: sd(x)
std_x <- sqrt(var_x)
# standard deviation of y: sd(y)
std_y <- sqrt(var_y)

# covariance of x and y: cov(x, y)
cov_x_y <- sum((x - x_bar)*(y - y_bar)) / (n - 1)
# correlation of x and y: cor(x, y)
cor_x_y <- cov_x_y / (std_x * std_y)

# estimate of predictor 1
beta_1 <- (sum((x - x_bar)*(y - y_bar))) / sum((x - x_bar)^2)
# estimate of intercept 
beta_0 <- y_bar - beta_1*x_bar
```

The current calculation for $\beta_1$ is the raw, simplified expression of dividing the covariance of x and y by the variation of x. Essentially the n - 1 terms cancel out. So, $\beta_1$ would also just equal `cov_x_y` / `var_x`. Algebraically, that expression is equivalent to `cor_x_y` * (`sd_y` / `sd_x`). So given seems to be the most computationally efficient. 

#### Calculation of $\hat{\sigma}$ (Residual Standard Error) 

```{r}
# number of predictors
p = 1

# degrees of freedom
dof <- n - p - 1

# predicted values of y from fitted equation
y_hat <- beta_0 + beta_1 * x
# residuals: actual - predicted values
residuals_y <- y - y_hat
# residual sum of squares
rss <- sum(residuals_y^2)

# residual standard error
sigma_hat <- sqrt(rss / dof)
```

The residual standard error, $\hat{\sigma}$ is the "typical" size of a residual. On average, the model's predictions are off by $\hat{\sigma}$

#### Calculation of Standard Error of $\beta_1$ and $\beta_0$

```{r}
# sums of squares of x
ss_x <- sum((x - x_bar)^2)

# standard error of beta_1
se_beta_1 <- sigma_hat / sqrt(ss_x)

# standard error of beta_0
se_beta_0 <- sigma_hat * sqrt((1 / n) + ((x_bar^2) / (ss_x)))
```

These values tell us that given a new sample from the population, how much does we think this estimate will deviate from the true population parameter

#### Calculation of t values and probabilities

Used for hypothesis testing the estimates of $\beta_p$

$$
	H_0: \beta_1 = 0 \qquad H_a: \beta_1 \neq 0
$$

```{r}
# the null value of the slope
h_0 <- 0

# t-value for x assuming the null is true
t_value_1 <- (beta_1 - h_0) / se_beta_1

# p-value for the t-value of x
p_value_1 <- 2 * pt(t_value_1, df=dof)

# t-value for intercept assuming the null is true
t_value_0 <- (beta_0 - h_0) / se_beta_0

# p-value for the t-value of the intercept
p_value_0 <- 2 * pt(abs(t_value_0), df=dof, lower.tail = FALSE)
```

While the code is different, the methods of obtaining the p-values are actually the same. This hypothesis test is a test of difference, which makes it a two sided test. The function `pt`, or the cumulative probability of the t-distribution, gives the area of the left. So then we multiply the result by two. We essentially do the same thing for the other p-value, but we multiply the result of he upper tail by 2. This works out since the distribution is symmetric. But in general, to get the correct p-value, use the left tail for negative values, and right tail for positive values due to floating point imprecision. 

#### Calculation of $R^2$

```{r}
# total sums of squares
tss <- sum((y - y_bar)^2)

# R^2
r_2 <- 1 - rss / tss
```

The total sum or squares is the total variance in the y variable, or how much the points vary around the mean. Residual sum of squares is the variance that is left over, or that is left unexplained by the model. Meaning that $R^2$ is the proportion of the total variance that the model explains. Or 1 minus the unexplained proportion, leaving the explained proportion.

#### Calculation of Adjusted $R^2$

```{r}
# mean square error (MSE)
mse <- rss / dof

# mean square total (MST)
mst <- tss / (n - 1)

# adjusted R^2
r_2_adj <- 1 - mse / mst

# alternate calculation when R^2 is already gotten
r_2_adj <- 1 - ((1 - r_2)*(n - 1)) / dof
```

#### Calculation of the F-statistic and p-value

```{r}
# sum of squares regression (SSR)
ssr <- sum((y_hat - y_bar)^2)

# mean square regression (MSR)
msr <- ssr / p

# F value
f_value = msr / mse

# p-value from F
f_p_value <- pf(f_value, p, dof, lower.tail = FALSE)
```

- TSS: How much the actual y values vary around the mean of y

- RSS: How much the actual y values vary around the predicted y values

- SSR: How much the predicted y values vary around the mean of y

TSS is your total variation. SSR is the part of that variation captured by the model's predictions ($\hat{y}$​), and RSS is the leftover "error" variation.

### Plots with Line of Best Fit

```{r}
plot(x = Boston$lstat, y = Boston$medv,
     xlab = "x (lstat)", ylab = "y (medv)",
     main = "Simple Linear Regression Fit",
     col="red", pch = 20, cex.lab = 1.5, cex.main = 1.5
)
abline(simple_model, col = "blue", lwd = 2)
```

```{r}
df |> 
	ggplot(aes(lstat, medv)) +
	geom_point(color="red") +
	geom_smooth(color="blue", se=FALSE, method="lm") +
	labs(
		title = "Simple Linear Regression Fit"
	)
```

## Multiple Linear Regression

### R Function

```{r}
multi_model <- lm(medv ~ lstat + rm + age, data=df)
summary(multi_model)
```

### Manual Calculation

#### Solving for $\hat{\beta}$

Standard matrix notation for solving

$$
	\hat{\beta} = (X^{\top}X)^{-1} X^{\top}Y
$$

```{r}
# matrix of all covariates
X <- model.matrix(medv ~ lstat + rm + age, data=df)

# response vector
Y <- df$medv

# X transpose X
XTX <- t(X) %*% X

# X transpose X inverse
XTX_inv <- solve(XTX)

# X transpose Y
XTY <- t(X) %*% Y

# plug in above for beta_hat
beta_hat <- XTX_inv %*% XTY
```

Alternative method that is a little more computationally efficient

$$
	(X^{\top}X)\hat{\beta} = X^{\top}Y
$$

```{r}
# solve(A, b) for x: Ax = b
solve(XTX, XTY)
```

#### Solving for Standard Errors

A few pieces need to solve this

$$
	\text{Var}(\hat{\beta}) = \sigma^2 (X^{\top}X)^{-1}
$$

Where $\sigma^2$ is going to be MSE, but we need

- RSS: needs $\hat{y}$

- DOF: degrees of freedom of the error

```{r}
P <- ncol(X) - 1

# predicted values
Y_hat <- X %*% beta_hat

# degrees of freedom for the error term
DoF <- n - P - 1

# residual sum of squares
RSS <- sum((Y - Y_hat)^2)

# a.k.a. sigma_hat^2
MSE <- RSS / DoF

# sigma_hat from MSE
standard_error <- sqrt(MSE)

# variance-covariance matrix of beta_hat
cov_matrix_beta <- MSE * XTX_inv

# variance of each beta
Var_beta <- diag(cov_matrix_beta)

# standard error of each beta
SE_beta <- as.matrix(sqrt(Var_beta))
```

#### Calculation of t values and probabilities

Used for hypothesis testing the estimates of $\beta_p$

$$
	H_0: \beta_p = 0 \qquad H_a: \beta_p \neq 0
$$

```{r}
# t-values of each beta estimate
t_values <- (beta_hat - h_0) / SE_beta

# p-value of each beta estimate, abs for consistency
p_values <- 2 * pt(abs(t_values), DoF, lower.tail = FALSE)
```

#### Calculation of $R^2$

```{r}
# same Y, so same TSS, but recalculate anyway
TSS <- sum((Y - mean(Y))^2)

# unadjusted R^2
R_2 <- 1 - RSS / TSS
```

#### Calculation of Adjusted $R^2$

```{r}
# using the alternate formula with R^2
R_2_adj <- 1 - ((1 - R_2)*(n - 1)) / DoF
```

#### Calculation of F-stat and probability

```{r}
# sun or squares regression
SSR <- sum((Y_hat - mean(Y))^2)

# mean square regression
MSR <- SSR / P

# Overall F-value
F_value <- MSR / MSE

# p-value from F
F_p_value <- pf(F_value, P, DoF, lower.tail = FALSE)
```

To get each predictors Sum Sq., each partial model would need to be fitted to do the subtraction process


































